---
title: "Getting started with dinarch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dinarch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(dinarch)
```

`dinarch` fits Dynamic Innovation Negative Binomial ARCH (NB-DINARCH) models
to count time series: an autoregressive negative-binomial model where the
mean depends on past counts and, optionally, covariates.

## Simulate some data

`dinarch_simulate()` generates a count series with a given autoregressive
coefficient `b`, dispersion `phi`, and (optionally) covariate effects
specified via a model formula.

```{r}
set.seed(2)
n <- 200
gdp <- rnorm(n)

sim <- dinarch_simulate(
  n = n, b = 0.3, phi = 8,
  formula = ~gdp, covariates = data.frame(gdp = gdp),
  beta = c(log(4), 0.5),
  seed = 2
)

dat <- data.frame(y = sim$y, index = sim$index, gdp = gdp)
head(dat)
```

## Fit the model

`dinarch_fit_ml()` fits the model by maximum likelihood. `formula` specifies
how covariates enter the mean (here, a single `gdp` effect plus an
intercept); `n_lags` sets the number of autoregressive lags.

```{r}
fit <- dinarch_fit_ml(dat, y = "y", index = "index", formula = ~gdp, n_lags = 1)
fit
summary(fit)
```

`summary()` reports standard errors and confidence intervals alongside the
point estimates.

## Predict forward

`predict()` draws samples from the fitted model's predictive distribution -
`horizon` sets how many periods ahead, `nsim` how many draws. Covariates are
held fixed at their last observed value by default.

```{r}
pred <- predict(fit, horizon = 10, nsim = 500, seed = 2)
head(pred)
```

`summary()` collapses the draws into a mean and prediction interval per
period; `plot()` shows the same thing as a shaded band (pass
`type = "spaghetti"` to instead draw every draw as its own thin line).

```{r, fig.width = 6, fig.height = 4}
summary(pred)
plot(pred)
```

Calling `predict()` with `horizon = NULL` (the default) instead replays the
fitted model in-sample, useful as a quick check that the fitted dynamics
look reasonable against the data the model was fit to.

`predict()` is a thin convenience wrapper around the more general
`dinarch_project()`, which additionally supports scenario-style
overrides (`new_data` for alternative covariate paths, `new_para` for
per-period parameter overrides) - see below.

## Scenario: transitioning between two fits

Suppose a second fit represents a different regime - `dinarch_transition_new_para()`
builds a `new_para` table that smoothly interpolates every parameter from
one fit to another over the forecast horizon, for scenarios like "assume
the dynamics gradually shift from `fit`'s regime to `fit_b`'s":

```{r, fig.width = 6, fig.height = 4}
sim_b <- dinarch_simulate(
  n = n, b = 0.7, phi = 8,
  formula = ~gdp, covariates = data.frame(gdp = gdp),
  beta = c(log(4), 1.2), seed = 3
)
dat_b <- data.frame(y = sim_b$y, index = sim_b$index, gdp = gdp)
fit_b <- dinarch_fit_ml(dat_b, y = "y", index = "index", formula = ~gdp, n_lags = 1)

new_para <- dinarch_transition_new_para(fit, fit_b, horizon = 10)
scenario <- dinarch_project(fit, horizon = 10, nsim = 500, new_para = new_para, seed = 2)
plot(scenario)
```

By the last forecast period, the simulated dynamics match `fit_b`'s
parameters exactly. `fit` and `fit_b` must share `n_lags` and the same
`beta` coefficient names (i.e. the same `formula`).

## Bayesian estimation

`dinarch_fit_bayes()` fits the same model via Stan/HMC instead of maximum
likelihood (requires the `rstan` package), and works as a drop-in
replacement for `dinarch_fit_ml()` in the calls above.
