---
title: "Getting started with dinarch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dinarch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(dinarch)
```

`dinarch` fits Dynamic Innovation Negative Binomial ARCH (NB-DINARCH) models
to count time series: an autoregressive negative-binomial model where the
mean depends on past counts and, optionally, covariates.

## Simulate some data

`dinarch_simulate()` generates a count series with a given autoregressive
coefficient `b`, dispersion `phi`, and (optionally) covariate effects
specified via a model formula.

```{r}
set.seed(2)
n <- 200
gdp <- rnorm(n)

sim <- dinarch_simulate(
  n = n, b = 0.3, phi = 8,
  formula = ~gdp, covariates = data.frame(gdp = gdp),
  beta = c(log(4), 0.5),
  seed = 2
)

dat <- data.frame(y = sim$y, index = sim$index, gdp = gdp)
head(dat)
```

## Fit the model

`dinarch_fit_ml()` fits the model by maximum likelihood. `formula` specifies
how covariates enter the mean (here, a single `gdp` effect plus an
intercept); `n_lags` sets the number of autoregressive lags.

```{r}
fit <- dinarch_fit_ml(dat, y = "y", index = "index", formula = ~gdp, n_lags = 1)
fit
summary(fit)
```

`summary()` reports standard errors and confidence intervals alongside the
point estimates.

## Forecast forward

`dinarch_project()` simulates forward from a fitted model - `horizon` sets
how many periods ahead, `nsim` how many stochastic replicates to draw.
Covariates are held fixed at their last observed value by default.

```{r}
forecast <- dinarch_project(fit, horizon = 10, nsim = 200, seed = 2)
head(forecast)

# median and a 90% interval per forecast period
agg <- forecast[, .(
  median = median(y),
  lo90 = quantile(y, 0.05),
  hi90 = quantile(y, 0.95)
), by = index]
agg
```

Calling `dinarch_project()` with `horizon = NULL` instead replays the fitted
model in-sample, which is useful as a quick check that the fitted dynamics
look reasonable against the data the model was fit to.

## Bayesian estimation

`dinarch_fit_bayes()` fits the same model via Stan/HMC instead of maximum
likelihood (requires the `rstan` package), and works as a drop-in
replacement for `dinarch_fit_ml()` in the calls above.
